On the Functionality of the Multilayer Perceptron

### 2020-12-05 Thesis/Drafts/Buell-PHYS799-ThesisDraft0.tex
Modified Figures, Modified Thesis

% ================================
% Landon Buell
% Kevin Short
% Physics 799%
% 26 September 2020
% ================================

@@

\begin{titlepage}
	\begin{center}
   		\vspace*{1cm}
       	\begin{LARGE}
       	\textbf{Mapping Chaotically Generated Synthesizers to Real World Instruments with Machine Learning and Harmonic Analysis} \\
       	\end{LARGE}
       	\vspace{1cm}
       	\begin{Large}
       	\textcolor{red}{Thesis Subtitle}
       	\end{Large}             
       	\vspace{1.5cm}

@@

@@

@@

\newpage
\section{Methodology}
\label{sec-Methodology}

@@

\subsection{Designing the Function}

@@

\subsection{Collecting and Pre-processing Raw Data}

@@

\subsection{Designing Classification Features}

@@

\subsection{Designing A Complementary Network Architecture}

@@

\subsection{Testing and Evaluating Network Performance}

@@

\subsection{Running Predictions of Chaotic Synthesizer Files}

@@

\paragraph*{}With this final change, we perform a \textit{Discrete Fourier Transform} (DFT) to bring each analysis frame from a time domain into a frequency domain \cite{Olson,Peatross,Virtanen}. The Discrete Fourier Transform is applied by producing an $N \times N$ \textit{transform matrix}, often noted at $\mathbb{W}$. Let $\omega^k = e^{\frac{-2\pi i}{N}k}$, then the DFT matrix for a time-space signal containing $N$ samples is given by \cite{Taylor,Virtanen}
\begin{equation}
\label{eqn-DFTMatrix}
\mathbb{W} = \frac{1}{\sqrt{N}}
\begin{bmatrix}
1 & 1 & 1 & 1 & \hdots & 1 \\
1 & \omega		& \omega^2 & \omega^3 & \hdots & \omega^{N-1} \\
1 & \omega^2	& \omega^4 & \omega^6 & \hdots & \omega^{2(N-1)} \\
1 & \omega^3	& \omega^6 & \omega^9 & \hdots & \omega^{3(N-1)} \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \omega^{N-1} & \omega^{2(N-1)} & \omega^{3(N-1)} & \hdots & \omega^{(N-1)^2}
\end{bmatrix}
\end{equation}
Each column of the matrix is a complex sinusoidal oscillating with an integer number of periods within the $N$-sample length window \cite{Short,Peatross}. The DFT is applied by a taking the matrix - product of $\mathbb{W}$ and $\widetilde{A}^T$. The transpose of $\widetilde{A}$ then makes each analysis frame into a column vector, which gives the appropriate dimension for multiplication.
\begin{equation}
\label{eqn-DFT}
\text{DFT}\big[ \widetilde{A} \big] = \mathbb{W}\widetilde{A}^T
\end{equation}

\paragraph*{}Most standard implementations of neural network models have all activations, weights, and biases to be real floating-point numbers cite{Levine,Loy,Tensorflow}. Since the elements of the DFT matrix lie on the complex unit circle, each real-valued analysis frame will be moved into complex space. This means we compute the element-wise product of the transformed signal matrix and it's complex-conjugate matrix. This is the $N \times k$
spectrogram matrix, where each element is a real-values number. Matrix $S$ is defined:
\begin{equation}
\label{eqn-Spectrogram}
S = (\mathbb{W}\widetilde{A}^T) \odot (\mathbb{W}\widetilde{A}^T)^*
\end{equation}
Where $\mathbb{W}$ is the DFT matrix from Eq. (\ref{eqn-DFTMatrix}) and $\widetilde{A}^T$ is the transpose of the analysis frames matrix from Eq. (\ref{eqn-WindowMatrix}). The asterisks indicate the element-wise complex conjugation. In practice, the matrix product $\mathbb{W}\widetilde{A}^T$ is computed once, and the conjugate is applied to a copy of the array.

@@

\paragraph*{}Standard western musical instruments seldom have fundamental frequencies that extend above $6$ kHz \cite{Olson,Virtanen,White}. This means that when constructing the spectrogram, we will rarely see significant energy present above this frequency at any time, and the $S$ matrix will contain mostly zero, or zero-like entries. To condense the size of the matrix, we select only the frequency bins (rows) that correspond to energies between $0$ Hz and $6000$ Hz. This makes the input array smaller, and eliminates redundant and non-useful information. The number rows in the $S$ matrix is reduced from $N$ down to $N'$.

\paragraph*{}Each spectrogram is now $N' \times k$ and effectively encodes the energy distribution of the waveform as a function of both time and frequency. The spectrogram makes up each sample in the first design matrix $X_1$, Eq.(\ref{eqn-ShapeX1}) used in this model. For this classifier, we have chosen $N' = 558$ and $k = 256$. For training, a batch of $b$ samples are concatenated into a single array object. For a batch of $b$ samples of $N' \times k \times 1$ spectrograms, we shape $X_1$ such that:
\begin{equation}
\label{eqn-X1}
X_1 = \big\{ S^{(0)},S^{(1)},S^{(2)}, ... , S^{(b-1)} \big\} \in \mathbb{R}^{(b \times N' \times k)}
\end{equation}
Which is consistent with the shape of the $X_1$ matrix outlined in eqn(\ref{eqn-X1Shape}). This matrix is presented to the \textit{Convolution} branch of the neural network for processing.

@@

\paragraph*{}The features described in this section are derived from time-domain representations of each audio sample. Each feature is one element in the feature vector for the particular sample. For consistency between samples, each waveform is padded or truncated to contain the same number of samples $M$. The number of samples $M$ is chosen to correspond the number of sample needed to make the $N \times k$ spectrogram matrix. This way,t he spectrogram and time-series features are representative of the same time interval in the file sample. Time space is indexed by $s[i]$ with $i \in [0,1,2,3,...,M-2,M-1]$

@@

\paragraph*{}One way to mitigate this is to group analysis frames together and compute the TDE of a collection of frames. This way, small most waveforms will produce an identical set of features to a time-delayed version of itself. By choosing a sufficiently large number of TDE values, we can also generate n approximation of than amplitude envelope of the time-series waveform. This comes at a higher computational cost, and a previously mentioned higher dimensional feature vector

\paragraph*{}We adapt this TDE to \textbf{compute the time domain envelope over $5$ non-overlapping analysis frames}. Since the maximum amplitude of each waveform array has been normalized to $\pm 1$, the TDE represents a consistent measurement of energy in each waveform subset. \cite{Liu}. The TDE is computed as the RMS-Energy of the waveform, $s$ and the $j$-th TDE value is given by \cite{Olson,Serizel}:
\begin{equation}
\label{eqn-RMS}
\text{TDE}_j\big[ s \big] = \sqrt{\frac{1}{Q} \sum_{i=n}^{n+Q} s[i]^2}
\end{equation}
Where $Q$ gives the number of sample sin each analysis frame (number of sample sin waveform divided by number of frames), and $n$ gives the index where the $j$-th frame begins.

\paragraph*{}These uniformly size analysis frames let us create an naive envelope of the waveform, but will allow it to generalize between classes. For example, this allows for a crude approximation of the energy in the attack, decay, sustain, and release energies in a waveform \cite{Virtanen,Olson}. For instruments with heavier attacks, we expect the TDE in the $1$-st frame to be comparably large, see Fig. (\ref{fig-FeatureTDE}a). Instruments with a heavy decay will likely have little energy in the $2$-nd or higher frames, indicating that the amplitude has substantially died off as in Fig. (\ref{fig-FeatureTDE}b). Conversely, instruments with longer sustain such as upper woodwinds, vibraphones or strings will container relatively higher TDE values for the later analysis frames as seen in Fig. (\ref{fig-FeatureTDE}c). Instruments with little or no waveform envelope, such as the synthesized waveforms and the whitenoise show constant TDE values across each frame as expected.

@@

\begin{figure}[H]
\begin{center}
%\textcolor{red}{PLOT OF ZXR FEATURES HERE}
\includegraphics[scale=0.3]{../FiguresFeatures/ZXR}
\end{center}
\caption{A comparison of the Zero-Crossing Rate for each class using a box-and-whisker plot}
\label{fig-FeatureZXR}
\end{figure}

@@

\paragraph*{}TCM condenses the idea of the amplitude envelope into a single value. It allows for the quick measurement of where in the time most of the energy of the signal lies. For percussive instruments with short attack and release times, such as bells or xylophones, we expect a very low TCM. Instruments with plucked strings, but longer release times such as mandolins and guitars should have similarly low values. Instruments in the woodwind and brass family generally have longer sustain and release times, giving them a slightly higher center of mass. Strings and undampened percussion have notoriously long sustain and release times giving them much higher TCM values \cite{Olson,White}. Finally, the synthetic wave forms have no characteristic  envelope shape (due to their synthetic nature). This gives a centrally located TCM, very close to $M/2$.

@@

\paragraph*{}\textcolor{red}{I need to complete my discussion of ACC's here!}

\begin{figure}[H]
\begin{center}
%\textcolor{red}{PLOT OF ACC FEATURES HERE}

@@

\end{center}
\caption{A comparison of the the first four auto correlation coefficents in each class using a box-and-whisker plot}
\label{fig-FeatureACC}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\textcolor{red}{Figure of a few Mel Filter Banks goes here}
\end{center}
\caption{Mel Filter Banks shown in frequency space with units of (a) Mels and (b)Hertz}
\label{fig-MelFilterBanks}
\end{figure}

@@

\paragraph*{}Physically, MFCC's are a transform of a transform. This allows us to investigate the periodicity of the frequency spectrum, which highlights phenomena such as overtones or echoes in the signal\cite{Virtanen}. Cepstrum coefficients are commonly used to speech identification and are very prolific in sound recognition tasks \cite{Serizel,Sahidullah,Liu}.

\paragraph*{}As the frequency bin index increases, the width of the bin in frequency space increases. \textcolor{red}{Finish this! What does a MFCC tell us with words!}

\begin{figure}[H]
\begin{center}
%\textcolor{red}{PLOT OF MFCC FEATURES HERE}

>>>>

% ================================
% Landon Buell
% Kevin Short
% Physics 799
% 26 September 2020
% ================================

@@

\begin{titlepage}
	\begin{center}
   		\vspace*{1cm}
       	\begin{LARGE}
       	\textbf{Mapping Chaotically Generated Synthesizers to Real World Instruments with Machine Learning and Harmonic Analysis} \\
       	\end{LARGE}
       	\vspace{1cm}
       	\begin{Large}
       	Multimodal Classification of Digital Signal
       	\end{Large}             
       	\vspace{1.5cm}

@@

\subsection{Introduction}

@@

\paragraph*{}Consider the task of identifying the source of a sound.

@@

\newpage
\subsection{Methodology}
\label{sec-Methodology}

@@

\subsubsection{Designing the Function}

@@

\subsubsection{Collecting and Pre-processing Raw Data}

@@

\subsubsection{Designing Classification Features}

@@

\subsubsection{Designing A Complementary Network Architecture}

@@

\subsubsection{Testing and Evaluating Network Performance}

@@

\subsubsection{Running Predictions of Chaotic Synthesizer Files}

@@

\paragraph*{}With this final change, we perform a \textit{Discrete Fourier Transform} (DFT) to bring each analysis frame from a time domain into a frequency domain \cite{Olson,Peatross,Virtanen}. The Discrete Fourier Transform is applied by producing an $N \times N$ \textit{transform matrix}, often noted at $\mathbb{W}$. Let $\omega^k = e^{\frac{-2\pi i}{N}k}$, then the DFT matrix for a time-space signal containing $N$ samples is given by \cite{Taylor,Virtanen}
\begin{equation}
\label{eqn-DFTMatrix}
\mathbb{W} = \frac{1}{N}
\begin{bmatrix}
1 & 1 & 1 & 1 & \hdots & 1 \\
1 & \omega		& \omega^2 & \omega^3 & \hdots & \omega^{N-1} \\
1 & \omega^2	& \omega^4 & \omega^6 & \hdots & \omega^{2(N-1)} \\
1 & \omega^3	& \omega^6 & \omega^9 & \hdots & \omega^{3(N-1)} \\
\vdots & \vdots & \vdots & \vdots & \ddots & \vdots \\
1 & \omega^{N-1} & \omega^{2(N-1)} & \omega^{3(N-1)} & \hdots & \omega^{(N-1)^2}
\end{bmatrix}
\end{equation}
Each column of the matrix is a complex sinusoidal oscillating with an integer number of periods within the $N$-sample length window \cite{Short,Peatross}. The DFT is applied by a taking the matrix - product of $\mathbb{W}$ and $\widetilde{A}^T$. The transpose of $\widetilde{A}$ then makes each analysis frame into a column vector, which gives the appropriate dimension for multiplication.
\begin{equation}
\label{eqn-DFT}
\text{DFT}\big[ \widetilde{A} \big] = \mathbb{W}\widetilde{A}^T
\end{equation}

\paragraph*{}Most standard implementations of neural network models have all activations, weights, and biases to be real floating-point numbers \cite{Levine,Loy,Tensorflow}. Since the elements of the DFT matrix lie on the complex unit circle, each real-valued analysis frame will be moved into complex space. This means we compute the element-wise product of the transformed signal matrix and it's complex-conjugate matrix. This is the $N \times k$
spectrogram matrix, where each element is a real-values number. Matrix $S$ is defined:
\begin{equation}
\label{eqn-Spectrogram}
S = (\mathbb{W}\widetilde{A}^T) \odot (\mathbb{W}\widetilde{A}^T)^*
\end{equation}
Where $\mathbb{W}$ is the DFT matrix from Eq. (\ref{eqn-DFTMatrix}) and $\widetilde{A}^T$ is the transpose of the analysis frames matrix from Eq. (\ref{eqn-WindowMatrix}). The asterisks indicate the element-wise complex conjugation. In practice, the matrix product $\mathbb{W}\widetilde{A}^T$ is computed once, and the conjugate is applied to a copy of the array.

@@

\paragraph*{}Standard western musical instruments seldom have fundamental frequencies that extend above $6$ kHz \cite{Olson,Virtanen,White}. This means that when constructing the spectrogram, we will rarely see significant energy present above this frequency at any time, and the $S$ matrix will contain mostly zero, or zero-like entries. To condense the size of the matrix, we select only the frequency bins (rows) that correspond to energies between $0$ Hz and $12,000$ Hz. This makes the input array smaller, and eliminates redundant and non-useful information. The number rows in the $S$ matrix is reduced from $N$ down to $N'$.

\paragraph*{}Each spectrogram is now $N' \times k$ and effectively encodes the energy distribution of the waveform as a function of both time and frequency. The spectrogram makes up each sample in the first design matrix $X_1$, Eq.(\ref{eqn-X1Shape}) used in this model. For this classifier, we have chosen $N' = 558$ and $k = 256$. For training, a batch of $b$ samples are concatenated into a single array object. For a batch of $b$ samples of $N' \times k \times 1$ spectrograms, we shape $X_1$ such that:
\begin{equation}
\label{eqn-X1}
X_1 = \big\{ S^{(0)},S^{(1)},S^{(2)}, ... , S^{(b-1)} \big\} \in \mathbb{R}^{(b \times N' \times k)}
\end{equation}
Which is consistent with the shape of the $X_1$ matrix outlined in Eq. (\ref{eqn-X1Shape}). This matrix is presented to the \textit{Convolution} branch of the neural network for processing.

@@

\paragraph*{}The features described in this section are derived from time-domain representations of each audio sample. Each feature is one element in the feature vector for the particular sample. For consistency between samples, each waveform is padded or truncated to contain the same number of samples $M$. The number of samples $M$ is chosen to correspond the number of sample needed to make the $N \times k$ spectrogram matrix. This way, the spectrogram and time-series features are representative of the same time interval in the file sample. Time space is indexed by $s[i]$ with $i \in [0,1,2,3,...,M-2,M-1]$

@@

\paragraph*{}One way to mitigate this is to group analysis frames together and compute the TDE of a collection of frames. This way,  almost all waveforms will produce an identical set of features to a short time-delayed version of itself. By choosing a sufficiently large number of TDE values, we can also generate an approximation of than amplitude envelope of the time-series waveform. This comes at a higher computational cost, and a previously mentioned higher dimensional feature vector

\paragraph*{}We adapt this TDE to \textbf{compute the time domain envelope over $5$ non-overlapping analysis frames}. Since the maximum amplitude of each waveform array has been normalized to $\pm 1$, the TDE represents a consistent measurement of energy in each waveform subset \cite{Liu}. The TDE is computed as the RMS-Energy of the waveform, $s$ and the $j$-th TDE value is given by \cite{Olson,Serizel}:
\begin{equation}
\label{eqn-RMS}
\text{TDE}_j\big[ s \big] = \sqrt{\frac{1}{Q} \sum_{i=n}^{n+Q} s[i]^2}
\end{equation}
Where $Q$ gives the number of sample sin each analysis frame (number of sample sin waveform divided by number of frames), and $n$ gives the index where the $j$-th frame begins. We provide a graphic representation of how the TDE compares to the amplitude of sections of a waveform in Fig. (\ref{fig-ExampleTDE}).

\begin{figure}[H]
	\begin{subfigure}[b]{0.45\textwidth}
	\begin{center}
	\includegraphics[width=1.0\linewidth]{../FiguresSpectrogram/GUITAR-B2_TDE}
	\caption{Guitar - B2}
	\end{center}
	\end{subfigure}	
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
	\begin{center}
	\includegraphics[width=1.0\linewidth]{../FiguresSpectrogram/Tuba-A2_TDE}
	\caption{Tuba - A2}
	\end{center}
	\end{subfigure}
	\caption{TDE Envelope values for musical instruments}
	\label{fig-ExampleTDE}
\end{figure}

\paragraph*{}The uniformly sized analysis frames let us create an naive envelope of the waveform, but will allow it to generalize between classes. For example, this allows for a crude approximation of the energy in the attack, decay, sustain, and release portion of the signal \cite{Virtanen,Olson}. For instruments with heavier attacks, we expect the TDE in the $1$-st frame to be comparably large, see Fig. (\ref{fig-FeatureTDE}a). Instruments with a heavy decay will likely have little energy in the $2$-nd or higher frames, indicating that the amplitude has substantially died off as in Fig. (\ref{fig-FeatureTDE}b). Conversely, instruments with longer sustain such as upper woodwinds, vibraphones or strings will container relatively higher TDE values for the later analysis frames as seen in Fig. (\ref{fig-FeatureTDE}c). Instruments with little or no waveform envelope, such as the synthesized waveforms and the whitenoise show constant TDE values across each frame as expected.

@@

\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.3]{../FiguresFeatures/ZXR}
\end{center}
\caption{A comparison of the Zero-Crossing Rate for each class using a box-and-whisker plot}
\label{fig-FeatureZXR}
\end{figure}

@@

\paragraph*{}TCM condenses the idea of the amplitude envelope into a single scalar value. It allows for the quick measurement of where in the time most of the energy of the signal lies. For percussive instruments with short attack and release times, such as bells or xylophones, we expect a very low TCM. Instruments with plucked strings, but longer release times such as mandolins and guitars should have similarly low values. See Fig. (\ref{fig-ExampleTCM}) for examples.
\begin{figure}[H]
	\begin{subfigure}[b]{0.45\textwidth}
	\begin{center}
	\includegraphics[width=1.0\linewidth]{../FiguresSpectrogram/Vibraphone-sustain-A5-R_TCM}
	\caption{Vibraphone - A5}
	\end{center}
	\end{subfigure}	
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
	\begin{center}
	\includegraphics[width=1.0\linewidth]{../FiguresSpectrogram/CELLO-F4_TCM}
	\caption{Cello - F4}
	\end{center}
	\end{subfigure}
	\caption{TCM values for musical instruments}
	\label{fig-ExampleTCM}
\end{figure}
Instruments in the woodwind and brass family generally have longer sustain and release times, giving them a slightly higher center of mass. Strings and undampened percussion have notoriously long sustain and release times giving them much higher TCM values \cite{Olson,White}. Finally, the synthetic wave forms have no characteristic  envelope shape (due to their synthetic nature). This gives a centrally located TCM, very close to $M/2$.

@@

\paragraph*{}Physically, the numerator of the ACC representing computing the dot product of the signal, $s[i]$ with a time-delayed version of itself $s[i+k]$, and  the denominator provides a normalization for the value. Dotting the signal and the time delay allows us to introduce a synthetic phase shift and compare the resulting relationship. If we chose $k$ to be equal, or similar to the number of samples that make-up a period or half-period of the waveform, then $s[i] \approx s[i+k]$ and then $\text{ACC}_k \rightarrow 1$. Alternatively, for many other values of $k$, frequent multiplication and summation of samples that are approximately zero result in $\text{ACC}_k << 1$. This make auto-correlation coefficients extremely useful for detecting periodicity in time-space \cite{Serizel}.

\begin{figure}[H]
\begin{center}

@@

\end{center}
\caption{A comparison of the the first four auto correlation coefficients in each class using a box-and-whisker plot}
\label{fig-FeatureACC}
\end{figure}

@@

\begin{figure}[H]
\begin{center}
\includegraphics[scale=0.28]{../FiguresSpectrogram/MelFilterBanks}
\end{center}
\caption{Mel Filter Banks shown in frequency space with units of Hertz}
\label{fig-MelFilterBanks}
\end{figure}

@@

\paragraph*{}Physically, MFCC's are a transform of a transform. This allows us to investigate the periodicity of the frequency spectrum, which highlights phenomena such as overtones or echoes in the signal \cite{Virtanen}. Cepstrum coefficients are commonly used to speech identification and are very prolific in sound recognition tasks \cite{Serizel,Sahidullah,Liu}. We show a feature-space representation of MFCC's for a selection of coefficients in Fig. (\ref{fig-FeatureMFCCs}).

\begin{figure}[H]
\begin{center}
